Language models as probabilistic cognitive models

### 2023-12-01 03-paper/paper-LLM-probmod.tex

The model constructed so far predicts probability zero for distractor choices, so that the human data shown in Figure~\ref{fig:refgame-counts}, where the distractor option was chosen in both conditions, would immediately rule out the model entirely.
It is therefore common to include a small error probability $\epsilon$, with which a choice would be made at random \citep[e.g.,][]{LeeWagenmakers2013:Bayesian-Cognit}, so that we get:
%
\begin{align*}
  P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}) = (1 - \epsilon_{c}) \  P_{r}(R_{l}, C; \alpha_{c}) +  \nicefrac{\epsilon_{c}}{3}\,,
\end{align*}
%
where $\epsilon_{c}$ is a (condition-specific) parameter giving the probability that a choice was made by randomly guessing.
\mf{FC objected to ``condition-specificity'' which is fair, but I leave it as it was; my reasoning: we use model checking to rule out models; in this setting the conservative strategy is to err on the side of making the models too powerful (even at the cost of conceptual plausibility)}

@@

Parameterized predictions, like in Equation~(\ref{eq:RSA-likelihood}), can be assessed in the light of the empirical data with the usual tools of Bayesian data analysis \citep[e.g.][]{GelmanCarlin2014:Bayesian-Data-A,McElreath2016:Statistical-Ret,Lambert2018:A-Students-Guid}.
Let $\alpha_{c}\sim \text{log-Normal}(0.5,1)$ have a reasonably wide log-Normal prior, and let $\epsilon_{c} \sim \text{Beta}(1,5)$ have a Beta prior favoring small values.
Using Stan \citep{Team2023:The-Stan-Core-L} for Bayesian inference, we obtain estimates of posterior credible values of model parameters (summary statistics of which are shown in Table~\ref{tab:sumStats}).\footnote{
  Posterior samples where generated for four chains with 2000 samples each, after a warm-up of 1000 samples. The ``adapt-delta'' value was set to 0.99. Converge was checked with $\hat{R}$-statistics \citep{GelmanRubin1992:Inference-from-}.
}

@@

  item-level  & RSA                & production     & 2.62 & 3.15 & 3.69 & 0.08 & 0.13 & 0.17 & 0.34 &  \\
  item-level  & RSA                & interpretation & 0.24 & 0.66 & 1.05 & 0.10 & 0.15 & 0.20 & 0.25 &  \\
  item-level  & LLM                & production     & 0.15 & 0.21 & 0.29 & 0.00 & 0.10 & 0.21 & 0.00 & * \\
  item-level  & LLM                & interpretation & 0.08 & 0.13 & 0.18 & 0.00 & 0.17 & 0.33 & 0.00 & * \\
  cond.-level & RSA                & production     & 2.59 & 3.16 & 3.71 & 0.08 & 0.13 & 0.18 & 0.49 &  \\
  cond.-level & RSA                & interpretation & 0.24 & 0.66 & 1.07 & 0.10 & 0.15 & 0.20 & 0.52 &  \\
  cond.-level & avg. scores        & production     & 0.19 & 0.22 & 0.26 & 0.00 & 0.04 & 0.11 & 0.05 & * \\
  cond.-level & avg. scores        & interpretation & 0.15 & 0.18 & 0.20 & 0.00 & 0.02 & 0.06 & 0.00 & * \\
  cond.-level & avg. probabilities & production     & 0.86 & 1.04 & 1.22 & 0.08 & 0.13 & 0.18 & 0.51 &  \\
  cond.-level & avg. probabilities & interpretation & 0.36 & 0.48 & 0.60 & 0.00 & 0.05 & 0.12 & 0.00 & * \\
  cond.-level & avg. WTA           & production     & 0.87 & 1.04 & 1.23 & 0.08 & 0.13 & 0.17 & 0.49 &  \\
  cond.-level & avg. WTA           & interpretation & 0.17 & 0.39 & 0.61 & 0.10 & 0.15 & 0.20 & 0.26 &  \\

@@

These results tell us that not all ways of deriving condition-level predictions by averaging over item-level variation are equally good.
Some approaches clearly fail basic checks for statistical goodness-of-fit.
On the positive side, we also find that there is at least one model with predictors based on LLM-measures, namely the \emph{average-WTA} model, which is able to recover the patterns in the data.
In other words, there is a way of deriving predictor values for condition-level forced-choice probabilities from an LLM such that, when fed into a common linking function (here with parameters $\alpha$ for optimization and $\epsilon$ for random error), the human choice probability can be reconstructed faithfully in its entirety.\footnote{
  A potential worry is that the \emph{average-WTA} model is trivial in the sense that it could have predicted \emph{any} data observation.
  Appendix~\ref{sec:range-prior-pred} shows that this is not the case.
  \mf{todo: redo these calculations with actual WTA!}
}
On the other hand, there is a stark contrast between the results of item-level and condition-level data.
The model that was able to properly fit the condition-level data builds on predictions for item-level choice that are clearly incompatible with the human item-level data.
The model seems to be ``right for the wrong reasons.''

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Range of prior predictions of intermediate-scope model}
\label{sec:range-prior-pred}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

It may seem that, given the freedom to adjust optimization $\alpha$ and random guessing rate $\epsilon$, some models might be able to explain \emph{any} data observation.
The intermediate-scope model, which was not discredited by model criticism in terms of recovery-based posterior predictive checks, is not trivial in this sense.
Figure~\ref{fig:prediction-range} shows the range of predictions that the intermediate-scope model is, in principle, capable of making (for the whole possible range of \(\alpha\) and \(\epsilon\) values, disregarding Bayesian priors).
There are distributions over response types which the model does not predict \emph{ex ante}.
For example, the model does not predict cases where the number of competitor choices is almost as high as the number of target choice, as would be the case if participants would consistently \emph{not} engage in pragmatic reasoning.

  \includegraphics[width=0.45\textwidth]{00-pics/prediction-range-prod.png}
  %
  \includegraphics[width=0.45\textwidth]{00-pics/prediction-range-inter.png}

  \caption{
    Range of predictions the intermediate-scope model makes for any pair of parameter values $\alpha_{c}$ and $\epsilon_{c}$.
    The light gray triangle with the red boundary is the probability simplex, i.e., the space of all possible 3-place probability vectors.
    The darker gray area in gray boundary lines is the subspace that is compatible with the ordering that the probability of choosing the target is bigger than that of the competitor, which in turn is bigger than that of the distractor.
    The blue area is the subspace of probabilistic predictions the intermediate-scope model makes under any value of its parameters.
    The orange dot in the middle is the ``Laplace point'' of equal probability for all three options, which the model predicts for $\alpha=0$ or $\epsilon=1$.
    The yellow dot is the ``vanilla prediction'' for $\alpha_{c}=1$ and $\epsilon_{c}=0$.
  }
  \label{fig:prediction-range}
\end{figure}

@@

>>>>

The model constructed so far predicts probability zero for distractor choices, so that the human data shown in Figure~\ref{fig:refgame-counts}, where the distractor option was chosen in both conditions, would immediately rule out the model entirely.
It is therefore common to include a small error probability $\epsilon$, with which a choice would be made at random \citep[e.g.,][]{LeeWagenmakers2013:Bayesian-Cognit}, so that we get:
%
\begin{align*}
  P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}) = (1 - \epsilon_{c}) \  P_{r}(R_{l}, C; \alpha_{c}) +  \nicefrac{\epsilon_{c}}{3}\,,
\end{align*}
%
where $\epsilon_{c}$ is a (condition-specific) parameter giving the probability that a choice was made by randomly guessing.\footnote{
  Since the RSA model predicts probability 0 for the distractor option, this model is, in principle, able to predict any probability distribution over the three choice categories that is compatible with the order: $P_{r}(R_{t}) \ge P_{r}(R_{c}) \ge P_{r}(R_{d})$.
  Intuitively, this is because with $\epsilon=0$, $P_{r}(R_{d}, C; \alpha_{c}) = 0$, so that there is an $\alpha$ for any ratio of predicted choice probabilities for target and competitor, as long as the target probability is no smaller than the competitor probability.
  The $\epsilon$-transformation is essentially a linear shift in the probability simplex towards the maximum entropy prediction, so that every prediction which obeys the ordering restriction above can be made for some pair of $\alpha$ and $\epsilon$.
  This prediction-triviality is met in two ways.
  For one, the Bayesian priors on model parameters soft-constrain the model, so that the \emph{ex ante} credible predictions do rule out many logically possible observations.
  For another, we break the triviality by assigning a non-zero probability to the prediction for the distractor option.
  The same triviality problem lurks for the \emph{average-WTA} model introduced in Section~\ref{llm-predictions-for-reference-games}, and the same solution is applied to it.
}

@@

Parameterized predictions, like in Equation~(\ref{eq:RSA-likelihood}), can be assessed in the light of the empirical data with the usual tools of Bayesian data analysis \citep[e.g.][]{GelmanCarlin2014:Bayesian-Data-A,McElreath2016:Statistical-Ret,Lambert2018:A-Students-Guid}.
Let $\alpha_{c}\sim \text{log-Normal}(1,1)$ have a reasonably wide log-Normal prior, and let $\epsilon_{c} \sim \text{Beta}(1,15)$ have a Beta prior favoring small values.
Using Stan \citep{Team2023:The-Stan-Core-L} for Bayesian inference, we obtain estimates of posterior credible values of model parameters (summary statistics of which are shown in Table~\ref{tab:sumStats}).\footnote{
  Posterior samples where generated for four chains with 2000 samples each, after a warm-up of 1000 samples. The ``adapt-delta'' value was set to 0.99. Converge was checked with $\hat{R}$-statistics \citep{GelmanRubin1992:Inference-from-}.
}

@@

  item-level  & RSA                & production     & 2.62 & 3.13 & 3.69 & 0.08 & 0.12 & 0.16 & 0.29 &  \\
  item-level  & RSA                & interpretation & 0.21 & 0.67 & 1.08 & 0.08 & 0.14 & 0.19 & 0.21 &  \\
  item-level  & LLM                & production     & 0.15 & 0.19 & 0.24 & 0.00 & 0.07 & 0.16 & 0.00 & * \\
  item-level  & LLM                & interpretation & 0.08 & 0.11 & 0.14 & 0.00 & 0.09 & 0.22 & 0.00 & * \\
  cond.-level & RSA                & production     & 2.62 & 3.14 & 3.70 & 0.08 & 0.12 & 0.17 & 0.50 &  \\
  cond.-level & RSA                & interpretation & 0.27 & 0.68 & 1.13 & 0.09 & 0.14 & 0.19 & 0.51 &  \\
  cond.-level & avg. scores        & production     & 0.19 & 0.22 & 0.25 & 0.00 & 0.03 & 0.08 & 0.05 &  \\
  cond.-level & avg. scores        & interpretation & 0.15 & 0.18 & 0.20 & 0.00 & 0.02 & 0.05 & 0.00 & * \\
  cond.-level & avg. probabilities & production     & 0.86 & 1.03 & 1.22 & 0.08 & 0.12 & 0.17 & 0.47 &  \\
  cond.-level & avg. probabilities & interpretation & 0.36 & 0.46 & 0.59 & 0.00 & 0.04 & 0.10 & 0.00 & * \\
  cond.-level & avg. WTA           & production     & 0.86 & 1.04 & 1.22 & 0.08 & 0.12 & 0.17 & 0.48 &  \\
  cond.-level & avg. WTA           & interpretation & 0.15 & 0.37 & 0.58 & 0.03 & 0.12 & 0.18 & 0.49 &  \\

@@

These results tell us that not all ways of deriving condition-level predictions by averaging over item-level variation are equally good.
Some approaches clearly fail basic checks for statistical goodness-of-fit.
On the positive side, we also find that there is at least one model with predictors based on LLM-measures, namely the \emph{average-WTA} model, which is able to recover the patterns in the data.
In other words, there is a way of deriving predictor values for condition-level forced-choice probabilities from an LLM such that, when fed into a common linking function (here with parameters $\alpha$ for optimization and $\epsilon$ for random error), the human choice probability can be reconstructed faithfully in its entirety.
On the other hand, there is a stark contrast between the results of item-level and condition-level data.
The model that was able to properly fit the condition-level data builds on predictions for item-level choice that are clearly incompatible with the human item-level data.
The model seems to be ``right for the wrong reasons.''

@@

The work presented here also contains considerable researcher degrees of freedom.
We only considered one out of several conceivable ways of carving a parameterized likelihood function from LLM-based predictors.
If future work would contribute to systematically exploring these, the main goal of this paper would have been met, which is to raise awareness for the possibility, perhaps even necessity, to scrutinize LLMs at the same level of rigorous detail as other models in cognitive science.

@@

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section{Range of prior predictions of intermediate-scope model}
% \label{sec:range-prior-pred}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% It may seem that, given the freedom to adjust optimization $\alpha$ and random guessing rate $\epsilon$, some models might be able to explain \emph{any} data observation.
% The intermediate-scope model, which was not discredited by model criticism in terms of recovery-based posterior predictive checks, is not trivial in this sense.
% Figure~\ref{fig:prediction-range} shows the range of predictions that the intermediate-scope model is, in principle, capable of making (for the whole possible range of \(\alpha\) and \(\epsilon\) values, disregarding Bayesian priors).
% There are distributions over response types which the model does not predict \emph{ex ante}.
% For example, the model does not predict cases where the number of competitor choices is almost as high as the number of target choice, as would be the case if participants would consistently \emph{not} engage in pragmatic reasoning.

% \begin{figure}[t]
%   \centering

%   \includegraphics[width=0.45\textwidth]{00-pics/prediction-range-prod.png}
%   %
%   \includegraphics[width=0.45\textwidth]{00-pics/prediction-range-inter.png}

%   \caption{
%     Range of predictions the intermediate-scope model makes for any pair of parameter values $\alpha_{c}$ and $\epsilon_{c}$.
%     The light gray triangle with the red boundary is the probability simplex, i.e., the space of all possible 3-place probability vectors.
%     The darker gray area in gray boundary lines is the subspace that is compatible with the ordering that the probability of choosing the target is bigger than that of the competitor, which in turn is bigger than that of the distractor.
%     The blue area is the subspace of probabilistic predictions the intermediate-scope model makes under any value of its parameters.
%     The orange dot in the middle is the ``Laplace point'' of equal probability for all three options, which the model predicts for $\alpha=0$ or $\epsilon=1$.
%     The yellow dot is the ``vanilla prediction'' for $\alpha_{c}=1$ and $\epsilon_{c}=0$.
%   }
%   \label{fig:prediction-range}
% \end{figure}

@@

   \hline
\end{tabular}
